---
title: "DATA607 Final Project Code"
author: "Miraj Patel"
date: "2025-05-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
install.packages("httr")
install.packages("jsonlite")
install.packages("tidyverse")
install.packages("ggplot2")

library(httr)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
```

Data set 1
```{r}
url <- "https://data.cdc.gov/api/views/fhky-rtsk/rows.json?accessType=DOWNLOAD"
response <- GET(url)
json_data <- content(response, "text")
vaccine_data_list <- fromJSON(json_data, flatten = TRUE)

#View the data
print(vaccine_data)

# Inspect the structure of the resulting list
str(vaccine_data_list)

# Check if the key containing the actual data is part of the list and convert it to a data frame
if ("data" %in% names(vaccine_data_list)) {
  vaccine_data <- as.data.frame(vaccine_data_list$data, stringsAsFactors = FALSE)
  
  # Get the column names
  column_names <- colnames(vaccine_data)

  # Print the column names
  print(column_names)
} else {
  print("The key containing the data was not found in the JSON structure")
}

```

Clean Data set 1
```{r}
#new dataframe of only the necessary columns
vaccine_data_filtered <- vaccine_data[, c("V9", "V12", "V13", "V14", "V15", "V16")]
new_column_names <- c("Vaccine Type", "Geography", "Birth.Year.Birth.Cohort", "Dimension.Type", "Dimension", "Estimate")
colnames(vaccine_data_filtered) <- new_column_names

#removing the second year in the column as it is not needed
vaccine_data_filtered$`Birth.Year.Birth.Cohort` <- sapply(vaccine_data_filtered$`Birth.Year.Birth.Cohort`, function(x) {
  sub("^(\\d{4})-\\d{4}$", "\\1", x)
})

#renaming for analysis
vaccine_data_filtered$'Vaccine Type'[vaccine_data_filtered$'Vaccine Type' == "≥1 Dose Varicella"] <- "Varicella"
vaccine_data_filtered$'Vaccine Type'[vaccine_data_filtered$'Vaccine Type' == "≥1 Dose MMR"] <- "Measles, Mumps, Rubella"
vaccine_data_filtered$'Vaccine Type'[vaccine_data_filtered$'Vaccine Type' == "Combined 7 Series"] <- "Combined Series"


#Combine all the counties into their states
vaccine_data_filtered$Geography <- ifelse(grepl("^TX", vaccine_data_filtered$Geography), "Texas", vaccine_data_filtered$Geography)
vaccine_data_filtered$Geography <- ifelse(grepl("^IL", vaccine_data_filtered$Geography), "Illinois", vaccine_data_filtered$Geography)
vaccine_data_filtered$Geography <- ifelse(grepl("^NY", vaccine_data_filtered$Geography), "New York", vaccine_data_filtered$Geography)
vaccine_data_filtered$Geography <- ifelse(grepl("^PA", vaccine_data_filtered$Geography), "Pennsylvania", vaccine_data_filtered$Geography)

clean_vaccine_data <- vaccine_data_filtered %>%
  filter(Dimension.Type == "Age") %>%
  mutate(
    Age_Value = case_when(
      grepl("Months", Dimension) ~ paste0(gsub("\\s*Months.*", "", Dimension), "m"),
      grepl("Days", Dimension) ~ paste0(gsub("\\D", "", Dimension), "d"),
      TRUE ~ Dimension
    ),
    Estimate = as.numeric(as.character(Estimate))
  ) %>%
  filter(!is.na(Estimate)) %>%
  group_by(`Vaccine Type`, Geography, Birth.Year.Birth.Cohort, Age_Value) %>%
  summarize(Estimate = mean(Estimate, na.rm = TRUE), .groups = "drop")

#Pivot wider to show better coverage of vaccines per age in states
clean_vaccine_data_wide <- clean_vaccine_data %>%
  pivot_wider(
    id_cols = c(`Vaccine Type`, Geography, Birth.Year.Birth.Cohort),
    names_from = Age_Value,
    values_from = Estimate
  )
```

Data set 2
```{r}
vaccine_data_2 <- read.csv('https://raw.githubusercontent.com/mirajpatel289/Data607/refs/heads/main/Vaccination%20Coverage%2019-35month%20.csv')
```

Clean Data set 2

```{r}
vaccine_data_2_filtered <- vaccine_data_2[, c("Year", "Vaccination", "Category", "Group", "Estimate" )]

clean_vaccine_data_2 <- vaccine_data_2_filtered %>%
  mutate(`Vaccination` = case_when(
    `Vaccination` %in% c(
      "Combined series (4:3:1:4:3:1:4)",
      "Combined series (4:3:1:3:3:1:4)",
      "Combined series (4:3:1:3:3:1)"
    ) ~ "Combined Series",
    Vaccination == "DTP/DT/DTaP (4 doses or more)" ~ "DTaP",
    Vaccination == "Polio (3 doses or more)" ~ "Polio",
    Vaccination %in% c("Hib (3 doses or more)", "Hib (primary series plus booster dose)") ~ "Hib",
    Vaccination == "Hepatitis B (3 doses or more)" ~ "Hep B",
    Vaccination == "PCV (4 doses or more)" ~ "PCV",
    TRUE ~ `Vaccination`
  ))

```

Graphs

```{r}
#Data Set 1
summarized_data_1 <- clean_vaccine_data %>%
  group_by(`Vaccine Type`, `Birth.Year.Birth.Cohort`) %>%
  summarize(Average_Estimate = mean(Estimate, na.rm = TRUE), .groups = "drop") %>%
  mutate(Dataset = "Dataset 1") #used for combination of the two data sets

#Bar graph
ggplot(summarized_data, aes(x = `Birth.Year.Birth.Cohort`, y = Average_Estimate, fill = `Vaccine Type`)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Average Vaccination Estimate by Birth Year/Cohort",
       x = "Birth Year/Cohort",
       y = "Average Estimate",
       fill = "Vaccine Type") +
  theme_minimal()

#Another plot, Heat Map, easier to tell the percentage than bar graph
ggplot(summarized_data, aes(x = `Birth.Year.Birth.Cohort`, y = `Vaccine Type`, fill = Average_Estimate)) +
  geom_tile() +
  labs(title = "Heatmap of Vaccination Estimates",
       x = "Birth Year/Cohort",
       y = "Vaccine Type",
       fill = "Average Estimate") +
  theme_minimal()


#Data Set 2
summarized_data_2 <- clean_vaccine_data_2 %>%
  group_by(Year, Vaccination) %>%
  summarize(Average_Estimate = mean(Estimate, na.rm = TRUE), .groups = "drop") %>%
  mutate(Dataset = "Dataset 2", `Birth.Year.Birth.Cohort` = Year, `Vaccine Type` = Vaccination) #used for combination of two data sets

#Bar Graph
ggplot(summarized_data_2, aes(x = Year, y = Average_Estimate, fill = Vaccination)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Average Vaccination Estimate by Year",
       x = "Year",
       y = "Average Estimate",
       fill = "Vaccination") +
  theme_minimal()

#Heat Map
ggplot(summarized_data_2, aes(x = Year, y = Vaccination, fill = Average_Estimate)) +
  geom_tile() +
  labs(title = "Heatmap of Vaccination Estimates",
       x = "Year",
       y = "Vaccination",
       fill = "Average Estimate") +
  theme_minimal()

### Combine two datasets

#Convert Birth.Year.Birth.Cohort to character in summarized_data_1
summarized_data_1 <- summarized_data_1 %>%
  mutate(`Birth.Year.Birth.Cohort` = as.character(`Birth.Year.Birth.Cohort`))

#Convert Birth.Year.Birth.Cohort to character in summarized_data_2
summarized_data_2 <- summarized_data_2 %>%
  mutate(`Birth.Year.Birth.Cohort` = as.character(Year),
         `Vaccine Type` = Vaccination)

combined_data <- bind_rows(summarized_data_1, summarized_data_2)

#Adjust the factor levels to ensure Dataset 2 appears before Dataset 1 so the years are in order
combined_data$Dataset <- factor(combined_data$Dataset, levels = c("Dataset 2", "Dataset 1"))

# Create the combined bar graph using combined data
ggplot(combined_data, aes(x = `Birth.Year.Birth.Cohort`, y = Average_Estimate, fill = `Vaccine Type`)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~Dataset, scales = "free_x") +
  labs(title = "Average Vaccination Estimate by Year/Cohort",
       x = "Year or Birth Year/Cohort",
       y = "Average Estimate",
       fill = "Vaccine Type") +
  theme_minimal()

# Create the combined heatmap using combined data
ggplot(combined_data, aes(x = `Birth.Year.Birth.Cohort`, y = `Vaccine Type`, fill = Average_Estimate)) +
  geom_tile() +
  facet_wrap(~Dataset, scales = "free_x") +
  labs(title = "Heatmap of Vaccination Estimates",
       x = "Year or Birth Year/Cohort",
       y = "Vaccine Type",
       fill = "Average Estimate") +
  theme_minimal()
```

Statistical Analysis

```{r}
#Correlation for Dataset 2
correlation_dataset_2 <- cor.test(as.numeric(clean_vaccine_data_2$Estimate), as.numeric(clean_vaccine_data_2$Year))
print(correlation_dataset_2)

#Correlation for Dataset 1
correlation_dataset_1 <- cor.test(as.numeric(clean_vaccine_data$Estimate), as.numeric(clean_vaccine_data$`Birth.Year.Birth.Cohort`))
print(correlation_dataset_1)

#Scatter plot for correlation visualization
ggplot(clean_vaccine_data_2, aes(x = Year, y = Estimate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Vaccination Estimate vs. Year (Dataset 2)",
       x = "Year",
       y = "Estimate",
       caption = paste("Correlation Coefficient:", round(correlation_dataset_2$estimate, 2))) +
  theme_minimal()

#Birth.Year.Birth.Cohort is numeric for proper model fitting
clean_vaccine_data$Birth.Year.Birth.Cohort <- as.numeric(clean_vaccine_data$Birth.Year.Birth.Cohort)

ggplot(clean_vaccine_data, aes(x = `Birth.Year.Birth.Cohort`, y = Estimate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Vaccination Estimate vs. Birth Year/Cohort (Dataset 1)",
       x = "Birth Year/Cohort",
       y = "Estimate",
       caption = paste("Correlation Coefficient:", round(correlation_dataset_1$estimate, 3))) +
  theme_minimal()
```

Data set 1: There is a statistically significant very weak positive correlation between vaccination estimates and birth year/birth cohort 2011 to 2021. This suggests that every new year, vaccination coverage shows a slight increase, although the correlation is very weak.

Data Set 2: There is a statistically significant weak negative correlation between vaccination estimates and year 1995 to 2009. This suggests that every new year, there might be a slight decrease in vaccination coverage, although this correlation is quite weak.

