---
title: "Homework 2"
author: "Miraj Patel"
date: "`r Sys.Date()`"
output: html_document
---
### Introduction

For this assignment, I asked my family members and my girlfriend to rate the 6 movies from 1 to 5. If they didn't see the movie, then I would include NA in that column.

## Connect to MySQL
Installing and loading packages to communicate with MySQL.

```{r load-packages, message=FALSE}

library(tidyverse)
library(DBI)
library(RMySQL)

con <- dbConnect(RMySQL::MySQL(),
                 dbname = "movie_ratings",
                 host = "localhost",
                 user = "root")
```

### Load Data into R dataframe

Loading the data from ratings database from MySQL into R dataframe. When working with data, we will not be fortunate to have all the data points/sets and there will be data missing. Just like in this case, not all of my family members watched all the movies. When making my table in SQL, I decided to leave the empty data sets as NULL to show a zero value. When importing to R, these values will show as NA.  
```{r code-chunk-label}
ratings_df <- dbGetQuery(con, "SELECT * FROM ratings")

print(ratings_df)
```

### Rearranging the table
Rearranging the table.
```{r}
ratings_table <- ratings_df %>%
  pivot_wider(names_from = Critic,
              values_from = Rating)

print(ratings_table)
```

### Close connection to MySQL
```{r, message =FALSE}
dbDisconnect(con)
```


...

